---
title: "Evaluating impact with ITS"
output: html_notebook
---


```{r setup}
# Load libraries
suppressPackageStartupMessages({
  library("readr")
  library("tidyr")
  library("dplyr")
  library("stringr")
  library("MASS")
  library("lubridate")
  library("ggplot2")
  library("gridExtra")
})

getwd()
```

## Introduction
We are going to analyze some data on pneumonia mortality from Ecuador. These data were drawn from a national mortality registers and were formatted into time series. The original paper describing these analyses by Oliveira and Shioda can be found: https://pubmed.ncbi.nlm.nih.gov/32448889/


## Read in the time series data
The data have already been formatted for your use. 

- J12_J18_prim is deaths with an underlying cause of death of pneumonia

- acm_noj_prim is number of deaths, excluding those with a respiratory cause (J chapter) listed


```{r}
ds1 <- readRDS( './Data/ec_2_59m.rds')
nndss_2016_2021 <- read_csv("Data/NNDSS.2016.2021.csv")
```

## Have a look at the data structure
```{r}
str(nndss_2016_2021)
```

```{r}
head(ds1)
```


## Specify dates of vaccine introduction
For all of these methods we need to know when the vaccine was introduced

```{r}
#YYYY-MM-DD
vax.intro.date <- as.Date('2010-08-01') #when introduce vaccine?

vax.eval.date <- as.Date('2011-08-01') #When start evaluation (usually at least 12 month after intro)

```


## Let's plot the time series
```{r, fig.width=8, fig.height=3}

p1 <- ggplot(ds1, aes(x=date, y=J12_J18_prim)) + #define x and y
      geom_line() + #Specify that you want a line plot
      ylab("J12-18, <1 yo") + #y axis label
      xlab("Date") +  #x axis label
      theme_classic() +  #Nice looking formayt
      ylim(0, NA) + #extend axis to 0
      geom_vline(xintercept=as.Date('2010-08-01'), col='red', lty=2) #reference line
   
   
p2 <- ggplot(ds1, aes(x=date, y=acm_noj_prim)) +
   geom_line() + #Specify that you want a line plot
      ylab("All-cause, <1 yo") + #y axis label
      xlab("Date") +  #x axis label
      theme_classic() +  #Nice looking formayt
      ylim(0, NA) + #extend axis to 0
      geom_vline(xintercept=as.Date('2010-08-01'), col='red', lty=2) #reference line

grid.arrange(p1,p2, nrow=1)
   
```

## Fitting a simple linear trend
Here the X variable is time and the y variable is deaths due to J12-18
```{r, fig.width=4, fig.height=4}

p1 <- ggplot(ds1, aes(x=date, y=J12_J18_prim)) +
      geom_point() +
      ylab("J12-18, <1 yo") +
      xlab("Date") +
      theme_classic() +
      theme(panel.spacing= unit(2,'lines') , axis.text.x=element_text(angle=90)) +
      geom_hline(yintercept=0, col='gray', lty=2) 
p1
```
We will use negative binomial regression with these data. For *count* (number of deaths per month), we use either Poisson or negative binomial regression. We could use a Poisson regression, but often the data will be "overdispersed", and as a result, confidence intervals will be incorrectly tight. You can address this by using either as negative binomial regression, or a quassipoisson model, where you model the amount of unexplained variation

```{r}

ds1$index <- 1:nrow(ds1)

   mod1 <- glm.nb(J12_J18_prim ~ index , data=ds1)
   summary(mod1)
```

extract the fitted values and plot them
```{r, fig.width=4, fig.height=4}

ds1.pred <- cbind.data.frame(ds1,'pred'=predict(mod1,type='response'))

p1a <- p1 +
      geom_line(color='#e41a1c', data=ds1.pred, aes(x=date, y=pred))
   
p1a

```
#Let's control for seasonality
Use a monthly 'dummy' variable that allows each month to have a different average
```{r}
ds1$month<-as.factor(month(ds1$date)) #create as a factor variable

ds1$month[1:36] #print first 36 observations
```
fit updated model
```{r}
mod1a <- glm.nb(J12_J18_prim~date +month, data=ds1)
summary(mod1a)
```


```{r}
ds1.pred1 <- cbind.data.frame(ds1.pred,'pred2'=predict(mod1a,type='response')) #extract fitted value and add to ds1.pred

p2 <- p1a +
      geom_line(color='#377eb8', data=ds1.pred1, aes(x=date, y=pred2))

p2


```

We might want to add an 'offset' variable that is used as a denominator. This could be population size; or all-non-respiratory causes of mortality 
```{r}
ds1$log.offset <- log(ds1$acm_noj_prim)
```

refit the model with an offset and look at results
```{r}
  mod2<-glm.nb(J12_J18_prim~ index +month+ offset(log.offset) , data=ds1)
   summary(mod2)
```

```{r}

ds1.pred2 <- cbind.data.frame(ds1.pred,'pred3'=predict(mod2,type='response')) #extract fitted value and add to ds1.pred

p3 <- p2+
      geom_line(color='#4daf4a', data=ds1.pred2, aes(x=date, y=pred3))

p3

```

## Now lets test if the trend or level of disease changes 

There are 3 ways we can do this. 
1. Interrupted time series with discontiguous segments. This fits different line segments through the data and tests whether the slope or level of disease changes. This can sometimes result in weird jumps when fitting the model

2. Interrupted time series with connected segments (Spline model), which allows the slope to change in the post-vaccine period but forces the change to be smooth

3. Fit the model to the pre-vaccine period only and extrapolate a trend to the post-vaccine period


## Method 1. Interrupted time series
First create 'dummy variables' that are 0 before introduction and 1 during the 
```{r}
ds1$period1 <- 0
ds1$period2 <- 0

ds1$period1[ds1$date>=vax.intro.date] <- 1

ds1$period2[ds1$date>=vax.eval.date] <- 1

ds1[,c('date','index','period1','period2')]
```

Ok, now let's fit our ITS model and look at results .

We will start with a simple step-change model to see if the average number of cases changes

```{r}
  mod3<-glm.nb(J12_J18_prim~index +month+ 
                 period1  +
                 period2 + #change in disease during evaluation period
                 offset(log.offset) , data=ds1)

summary(mod3)
```

```{r}
ds.pred.its <- cbind.data.frame(ds1,'pred4'=predict(mod3,type='response')) #extract fitted value and add to ds1.pred

p4 <- ggplot(ds.pred.its, aes(x=date, y=J12_J18_prim)) +
      geom_point() +
      ylab("J12-18, <1 yo") +
      xlab("Date") +
      theme_classic() +
      theme(panel.spacing= unit(2,'lines') , axis.text.x=element_text(angle=90)) +
      geom_hline(yintercept=0, col='gray', lty=2) +
      geom_line(color='#377eb8', data=ds.pred.its, aes(x=date, y=pred4))
p4   
```


Now get an estimate for the counterfactual (ie hold vaccine effect to 0 in the model)

```{r}
#set the vacccine effect variables to 0
ds1.counterfactual<-ds1
ds1.counterfactual$period1<-0
ds1.counterfactual$period2<-0

#Generate the fitted values
ds.pred.its$pred4.cf <- predict(mod3,type='response',newdata=ds1.counterfactual) #extract fitted value and add to ds1.pred

#overlay onto plot
p4 +
 geom_line(color='#e41a1c', lty=2,data=ds.pred.its, aes(x=date, y=pred4.cf))


```


Calculate a rate ratio from this model as the ratio of the fitted value and the counterfactual values. We won't try to estimate a confidence interval here. The InterventionEvaluatR package does this for us automatically.
```{r}
ds.pred.its$rr.its <- ds.pred.its$pred4/ds.pred.its$pred4.cf

p6<- ggplot(ds.pred.its, aes(x=date, y=rr.its)) +
      geom_line() +
      ylab("Rate ratio") +
      xlab("Date") +
       labs(title='Rate Ratio ITS model')+
      theme_classic() +
      theme(panel.spacing= unit(2,'lines') , axis.text.x=element_text(angle=90)) +
      geom_hline(yintercept=1, col='gray', lty=2) 

p6
```


We could also try a slightly more complicated model where the slope of the line is also allowed to vary in the post vaccine period. This shows that the AIC score is worse (higher score) than the model that just accounts for a step change; so we will prefer that simpler model.

```{r}
  mod3a <- glm.nb(J12_J18_prim~index +month+ 
                 period1  + period1*index +
                 period2 + period2*index + #change in disease during evaluation period
                 offset(log.offset) , data=ds1)

 summary( mod3a)
AIC(mod3, mod3a)
```


## Approach 2: ITS Spline model

In this case, we force the changes to be smooth so that we don't get the weird jumps like in the ITS example

```{r}
intro.index <- which(ds1$date==vax.intro.date)
eval.index <- which(ds1$date==vax.eval.date)

ds1$spl1<-ds1$index - intro.index +1
ds1$spl2<-ds1$index - eval.index +1

ds1$spl1[ds1$spl1<0]<-0 
ds1$spl2[ds1$spl2<0]<-0 

ds1[65:80,c('date','index','spl1','spl2')]
```

Ok, now let's fit our spline model and look at results 
```{r}
  mod4<-glm.nb(J12_J18_prim~index +month+ 
                 index+ #trend pre-vaccine
                 spl1+spl2+ #post-vaccine changes
                 offset(log.offset) , data=ds1)
summary(mod4)

#AIC(mod3,mod4)
```

Get fitted and counterfactual for the spline model 
```{r}
 ds.pred.spl<-cbind.data.frame( ds1,'pred.spl'=predict(mod4, type='response') )

ds1.counterfactual.spl<-ds1
ds1.counterfactual.spl$spl1<-0
ds1.counterfactual.spl$spl2<-0

ds.pred.spl$pred.spl.cf <-predict(mod4, type='response', newdata=ds1.counterfactual.spl)

```

The red is counterfactual, blue is fitted
```{r}

p7 <- ggplot(ds.pred.spl, aes(x=date, y=J12_J18_prim)) +
      geom_point() +
      ylab("J12-18, <1 yo") +
      xlab("Date") +
      theme_classic() +
      theme(panel.spacing= unit(2,'lines') , axis.text.x=element_text(angle=90)) +
      geom_hline(yintercept=0, col='gray', lty=2) +
      geom_line(color='#377eb8', data=ds.pred.spl, aes(x=date, y=pred.spl)) +
       geom_line(color='#e41a1c', lty=2,data=ds.pred.spl, aes(x=date, y=pred.spl.cf))
p7

```


We can see here that the decline follows a smoother trajectory, by design

```{r}
ds.pred.spl$rr.spline <- ds.pred.spl$pred.spl/ds.pred.spl$pred.spl.cf

p8 <- p6 +
     geom_line(color='red', lty=2,data=ds.pred.spl, aes(x=date, y=rr.spline))
   
p8



```

## Approach #3: Fit to pre-vaccine period and extrapolate

First set the pneumonia variable to NA during the post-vaccine period
```{r}
ds1$J12_J18_prim_pre <- ds1$J12_J18_prim

ds1$J12_J18_prim_pre[which(ds1$date>=vax.intro.date)] <- NA

ds1[,c('date','J12_J18_prim_pre' )]
```


```{r}
mod5<- glm.nb(J12_J18_prim_pre~  index #time trend
              +month #seasonality
              +offset(log.offset), data=ds1)
```

```{r}
ds.pred.mod5<-cbind.data.frame(ds1, 'pred.mod5'=predict(mod5,type='response', newdata=ds1))
```

Plot the observed and extrapolated data. Note that here the comparison is the *observed* data with the model estimate extrapolated from the pre-vaccine period. This is different form the previous models where we were just comparing the model fitted value with the model fitted value with the vaccine-effects held to 0.
```{r}
p9 <- ggplot(ds.pred.mod5, aes(x=date, y=J12_J18_prim)) +
      geom_point() +
      ylab("J12-18, <1 yo") +
      xlab("Date") +
      theme_classic() +
      theme(panel.spacing= unit(2,'lines') , axis.text.x=element_text(angle=90)) +
      geom_hline(yintercept=0, col='gray', lty=2) +
      geom_line(color='#377eb8', data=ds.pred.mod5, aes(x=date, y=pred.mod5)) 
p9

```

```{r}
ds.pred.mod5$rr.trend<-ds1$J12_J18_prim/ds.pred.mod5$pred.mod5

```

##Let's compare the rate ratio estimates from the 3 models

```{r}
p8 +  geom_line(color='#377eb8',lwd=1, lty=1,data=ds.pred.mod5, aes(x=date, y=rr.trend))

```


## look at rate ratio during the evaluation period
As a rough estimate of this, let's just average the point-by-point estimates of the rate ratio during the evaluation period.

```{r}
eval.period<-ds1$date> vax.eval.date

rr.its.eval<- mean(ds.pred.its$rr.its[eval.period])

rr.spline.eval<- mean(ds.pred.spl$rr.spline[eval.period])

rr.trend.eval<- sum(ds.pred.mod5$J12_J18_prim[eval.period])/sum(ds.pred.mod5$pred.mod5[eval.period])

round(c('ITS'=rr.its.eval,'Spline'=rr.spline.eval,'Trend Extrapolation'=rr.trend.eval ),2)

#Percent decline
100*(1- round(c('ITS'=rr.its.eval,'Spline'=rr.spline.eval,'Trend Extrapolation'=rr.trend.eval ),2))

```

*Depending on the dataset, these 3 methods might or might not agree*













